---
title: "Forecasting Ecuador’s Oil Production: Assessing the impact of halting exploitation in Block 43-ITT"
author: "Leines Nicole & Martinez Sayra"
date: "2025-04-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,          # show code; hide if not needed
  message = FALSE,
  warning = FALSE
)
```

## Packages

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(writexl)
library(ggplotify)
library(janitor)
```

## Directory

```{r}
base_dir <- "D:/Geani/Box/Home Folder gnl13/Private/1 Academics/3 Time series/LeinesMartinez_ENV797_TSA_FinalProject" # Update this
data_dir <- file.path(base_dir, "Data")
output_dir <- file.path(base_dir, "Output")
images_dir <-file.path(base_dir, "Images")

file1 <- "oil_production_EC_2007_2024.xlsx"
file2 <- "oil_price_2007-2024.xlsx"
file3 <- "annual_oil_production_1972-2023.xlsx"
file4 <- "oil_data_2007_2024.xlsx"

file_path1 <- file.path(data_dir, file1)
file_path2 <- file.path(data_dir, file2)
file_path3 <- file.path(data_dir, file3)
file_path4 <- file.path(data_dir, file4)

oil_production <- read_excel(file_path1) %>% clean_names()
oil_prices <- read_excel(file_path2) %>% clean_names()
annual_oil_production <- read_excel(file_path3) %>% clean_names()
oil_data <- read_excel(file_path4) %>% clean_names()

```

## Introduction

Ecuador's economy has been heavily reliant on oil exploitation for over five decades. As a result, the oil revenue is the most important driver of the national GDP (García-Alban et al., 202).


```{r}
image_path1 <- file.path(images_dir, "Image1.png")
knitr::include_graphics(image_path1)

```


## Motivation

## Relevance

## Objectives

## Dataset information

## Analysis (Methods and Models)



## Wrangling data - Aggregate the hourly data to daily using averages

```{r}

#Demand

# from wide (h1–h24) to long format, convert hour to integer
load_long <- load_raw %>%
  pivot_longer(
    cols        = starts_with("h"),
    names_to    = "hour",
    names_prefix= "h",
    values_to   = "load_kwh"
  ) %>%
  mutate(
    date     = as_date(date),
    hour     = as.integer(hour),
    meter_id = factor(meter_id)
  )

# compute per‐meter daily mean
daily_load <- load_long %>%
  group_by(meter_id, date) %>%
  summarise(
    daily_load_kwh = mean(load_kwh, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
# Temperature 

# daily mean across all hours and stations
daily_temp <- temp_raw %>%
  pivot_longer(
    cols         = starts_with("t_ws"),
    names_to     = "station",
    names_prefix = "t_ws",
    values_to    = "temp_c"
  ) %>%
  mutate(date = as_date(date)) %>%
  group_by(date) %>%
  summarise(
    mean_temp_c = mean(temp_c, na.rm = TRUE),
    .groups     = "drop"
  )

# Relative humidity
daily_hum <- hum_raw %>%
  pivot_longer(
    cols         = starts_with("rh_ws"),
    names_to     = "station",
    names_prefix = "rh_ws",
    values_to    = "rh_pct"
  ) %>%
  mutate(date = as_date(date)) %>%
  group_by(date) %>%
  summarise(
    mean_rh_pct = mean(rh_pct, na.rm = TRUE),
    .groups     = "drop"
  )


```

```{r}

daily_data <- daily_load %>%
  inner_join(daily_temp, by = "date") %>%
  inner_join(daily_hum,  by = "date")

# total system demand per day, plus averaged covariates
agg_daily <- daily_data %>%
  group_by(date) %>%
  summarise(
    demand_kwh = sum(daily_load_kwh, na.rm = TRUE),
    temp_c     = mean(mean_temp_c,  na.rm = TRUE),
    rh_pct     = mean(mean_rh_pct,  na.rm = TRUE),
    .groups    = "drop"
  )

```

```{r}

agg_daily %>%
  slice_head(n = 6) %>%
  kable(
    caption = "First six days of aggregated daily demand, temperature, and humidity",
    digits  = 2
  ) %>%
  kable_styling(full_width = FALSE)

```

## Define training and testing sets

```{r}


train <- agg_daily %>%
  filter(date >= ymd("2005-01-01") & date <= ymd("2009-12-31"))

test <- agg_daily %>%
  filter(date >= ymd("2010-01-01") & date <= ymd("2010-02-28"))

full_train <- agg_daily %>%
  filter(date >= ymd("2005-01-01") & date <= ymd("2010-12-31"))

# 5.2 Construct msts objects with weekly and yearly seasonality
y_train <- msts(
  train$demand_kwh,
  seasonal.periods = c(7, 365.25),
  start             = c(2005, 1)
)

y_test <- msts(
  test$demand_kwh,
  seasonal.periods = c(7, 365.25),
  start             = c(2010, 1)
)

y_full <-  msts(
  full_train$demand_kwh,
  seasonal.periods = c(7, 365.25),
  start             = c(2005, 1)
)

```

## Forecasting until 2010

```{r}

# 1) Forecast horizon
h <- length(y_test)

# 2) Fit each model

# ARIMA + Fourier + weather regressors
f_tr          <- fourier(y_train, K = c(2,6))
f_ts          <- fourier(y_train, K = c(2,6), h = h)
fit_reg       <- auto.arima(
                   y_train,
                   seasonal = FALSE,
                   xreg     = cbind(f_tr, train$temp_c, train$rh_pct)
                 )
fc_reg        <- forecast(
                   fit_reg,
                   h    = h,
                   xreg = cbind(f_ts, test$temp_c, test$rh_pct)
                 )        


# ETS with Box–Cox (lambda chosen to stabilize MAPE)
fc_ets_bc     <- forecast(ets(y_train, lambda="auto"), h = h)

fc_naive       <- naive(y_train, h = h)
fc_tbats       <- forecast(tbats(y_train),   h = h)

# TBATS with weather covariates
xreg_temp      <- matrix(train$temp_c,  ncol = 1)
xreg_test_temp <- matrix(test$temp_c,   ncol = 1)
xreg_hum       <- matrix(train$rh_pct,  ncol = 1)
xreg_test_hum  <- matrix(test$rh_pct,   ncol = 1)
xreg_both      <- cbind(train$temp_c,  train$rh_pct)
xreg_test_both <- cbind(test$temp_c,   test$rh_pct)

fc_tbats_temp  <- forecast(tbats(y_train, xreg = xreg_temp),  h = h, xreg = xreg_test_temp)
fc_tbats_hum   <- forecast(tbats(y_train, xreg = xreg_hum),   h = h, xreg = xreg_test_hum)
fc_tbats_both  <- forecast(tbats(y_train, xreg = xreg_both),  h = h, xreg = xreg_test_both)


```

```{r results='asis'}
# 3) Extract accuracy metrics
models <- list(
    `ARIMA+F+Wx`   = fc_reg,
    `ETS-BoxCox`   = fc_ets_bc,
    Naive          = fc_naive,
    TBATS          = fc_tbats,
  `TBATS+Temp`   = fc_tbats_temp,
  `TBATS+Hum`    = fc_tbats_hum
)

accuracy_tbl <- purrr::map_df(models, function(fit) {
  acc <- accuracy(fit, y_test)
  # if accuracy() returned two rows (train & test), use the test row
  idx <- if (nrow(acc) == 2) 2 else 1
  # pull out the six metrics we want
  acc[idx, c("ME","RMSE","MAE","MAPE","MASE","ACF1"), drop=FALSE] %>%
    as_tibble(rownames = "dummy") %>%
    select(-dummy)
}, .id = "Model")

# 4) Render comparison table, sorted by MAPE
accuracy_tbl %>%
  arrange(MAPE) %>%
  kable(
    caption = "Forecast accuracy comparison (all models)",
    digits  = 2
  ) %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#F0F0F0")  # highlight best MAPE
```

## Forecasting until 2011- top 5 models - with all information

```{r}

forecast_dates <- seq(as.Date("2011-01-01"), as.Date("2011-02-28"), by = "day")
h_future       <- length(forecast_dates)  # 59

```

```{r}
#Model 1: # ETS with Box–Cox (lambda chosen to stabilize MAPE)
fc_ets_bc_full     <- forecast(ets(y_full, lambda="auto"), h = h_future)

print(fc_ets_bc_full)
plot(fc_ets_bc_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_ets_bc_full, series="ETS with Box–Cox",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_ets_bc_full)
```

```{r}
# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_ets_bc_full$mean)
  )
out_name <- "submission_final_ETS_with_Box–Cox.csv"
readr::write_csv(submission, file.path(output_dir, out_name))
```

```{r}
#Model 2: Naive

fc_naive_full <- naive(y_full, h = h_future)
print(fc_naive_full)
plot(fc_naive_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_naive_full, series="Naive",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_naive_full)

```

```{r}

# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_naive_full$mean)
  )
out_name <- "submission_final_Naive.csv"
readr::write_csv(submission, file.path(output_dir, out_name))

```

```{r}
#Model 3: TBATS

fc_tbats_full       <- forecast(tbats(y_full),   h = h_future)
print(fc_tbats_full)
plot(fc_tbats_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_tbats_full, series="TBATS",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_tbats_full)


```

```{r}

# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_tbats_full$mean)
  )
out_name <- "submission_final_TBATS.csv"
readr::write_csv(submission, file.path(output_dir, out_name))

```

```{r}


#Model 4-6: TBATS + regressors


#–– 1) Forecast temp & humidity into 2011 (no drift/mean) ––#
fit_temp <- auto.arima(full_train$temp_c, seasonal=FALSE,
                       allowdrift=FALSE, allowmean=FALSE)
fit_hum  <- auto.arima(full_train$rh_pct, seasonal=FALSE,
                       allowdrift=FALSE, allowmean=FALSE)

fc_temp <- forecast(fit_temp, h = h_future)$mean
fc_hum  <- forecast(fit_hum,  h = h_future)$mean

#–– 2) Build xreg matrices – just the covariates ––#
# Historical (2005–2010):
xreg_full_temp <- matrix(full_train$temp_c, ncol=1)
xreg_full_hum  <- matrix(full_train$rh_pct, ncol=1)
xreg_full_both <- cbind(full_train$temp_c, full_train$rh_pct)

# Future (2011):
xreg_future_temp <- matrix(fc_temp, ncol=1)
xreg_future_hum  <- matrix(fc_hum,  ncol=1)
xreg_future_both <- cbind(as.numeric(fc_temp), as.numeric(fc_hum))







```

```{r}

# Model 4: TBATS + Temp 
fit_tb_temp_full   <- tbats(y_full, xreg = xreg_full_temp)
fc_tbats_temp_full <- forecast(fit_tb_temp_full,
                          h    = h_future,
                          xreg = xreg_future_temp)

print(fc_tbats_temp_full)
plot(fc_tbats_temp_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_tbats_temp_full, series="TBATS",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_tbats_temp_full)
```

```{r}

# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_tbats_temp_full$mean)
  )
out_name <- "submission_final_TBATS_temp.csv"
readr::write_csv(submission, file.path(output_dir, out_name))

```

```{r}

# Model TBATS + Hum 
fit_tb_hum_full   <- tbats(y_full, xreg = xreg_full_hum)
fc_tbats_hum_full <- forecast(fit_tb_hum_full,
                         h    = h_future,
                         xreg = xreg_future_hum)

print(fc_tbats_hum_full)
plot(fc_tbats_hum_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_tbats_hum_full, series="TBATS",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_tbats_hum_full)
```

```{r}

# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_tbats_hum_full$mean)
  )
out_name <- "submission_final_TBATS_hum.csv"
readr::write_csv(submission, file.path(output_dir, out_name))

```

```{r}

# Model 6: TBATS + Both
fit_tb_both_full   <- tbats(y_full, xreg = xreg_full_both)
fc_tbats_both_full <- forecast(fit_tb_both_full,
                          h    = h_future,
                          xreg = xreg_future_both)

print(fc_tbats_both_full)
plot(fc_tbats_both_full)

#Plot model + observed data
autoplot(y_full) +
  autolayer(fc_tbats_both_full, series="TBATS",PI=FALSE) +
  ylab("Electricity demand")


checkresiduals(fc_tbats_both_full)
```

```{r}

# Export to CSV 
submission <- template %>%
  mutate(
    date = format(forecast_dates, "%Y-%m-%d"),
    load = as.numeric(fc_tbats_both_full$mean)
  )
out_name <- "submission_final_TBATS_both.csv"
readr::write_csv(submission, file.path(output_dir, out_name))

```

## Graph

```{r}
autoplot(y_full) +
  autolayer(fc_ets_bc_full, PI=FALSE, series="ETS") +
  autolayer(fc_naive_full, PI=FALSE, series="NAIVE") +
  autolayer(fc_tbats_full,PI=FALSE, series="TBATS") +
  autolayer(fc_tbats_temp_full,PI=FALSE, series="TBATS+T") +
  autolayer(fc_tbats_hum_full,PI=FALSE, series="TBATS+H") +
  autolayer(fc_tbats_both_full,PI=FALSE, series="TBATS+BOTH") +
  xlab("Day") + ylab("Daily Electricity demand") +
  guides(colour=guide_legend(title="Forecast"))
```

### Summary and Conclusions
